---
title: "Fate of soil nitrogen under long-term bare fallow - supplementry"
author: "Trish Fraser and Tom Moore"
date: last-modified
output-file: "Analysis_Report.html"
format:
  html:
    toc: true
    toc-depth: 5
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    page-layout: full
    max-width: 2000px
    fig-width: 8
    fig-height: 4
    code-tools: true
    code-fold: true
    code-overflow: scroll
    df-print: paged
    embed-resources: true
    css: css/aicc-tables.css
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| results: false

# CRAN mirror for non-interactive installs
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install and load packages
packages <- c(
  "tidyverse", "readxl", "writexl", "here", "janitor",
  "lme4", "MuMIn", "performance",
  "RColorBrewer", "gridExtra", "colorspace"
)
installed <- rownames(installed.packages())
missing   <- setdiff(packages, installed)
if (length(missing) > 0) install.packages(missing, dependencies = TRUE)
for (p in packages) library(p, character.only = TRUE)

options(mc.cores = parallel::detectCores())
```

# Data summary

::: {.callout-note title="Wrangling & Summary" collapse="true"}
```{r}
#| label: wrangle-df

# Total soil C & N: read, clean names, standardise depth labels and factors
df <- read_excel(here("Analysis/Data/compiled_data_for_paper.xlsx"), sheet = "Total soil C & N")
orig_names_1 <- names(df)
df <- df %>% clean_names()
response_labels_1 <- setNames(orig_names_1, names(df))
response_labels_1["tc_g_kg_soil"] <- "Total soil C (kg/ha)"
response_labels_1["tn_g_kg_soil"] <- "Total soil N (kg/ha)"

df <- df %>%
  mutate(depth = fct_recode(depth, "0-75" = "0-7.5 cm", "75-150" = "7.5-15 cm", "150-250" = "15-25 cm"))
cols <- c("trial", "year", "plot", "treatment", "rep", "depth")
df[, cols] <- lapply(df[, cols], as.factor)
if ("treat_ment" %in% names(df) && !"treatment" %in% names(df)) {
  df <- df %>% mutate(treatment = treat_ment) %>% dplyr::select(-treat_ment)
}
df %>% group_by(year, treatment, depth) %>% tally()
```

```{r}
#| label: wrangle-df2

# POM C & N: read (skip header), clean names, depth labels, factors; ensure "months" for modelling
df.2 <- read_excel(here("Analysis/Data/compiled_data_for_paper.xlsx"), sheet = "POM C & N", skip = 1)
orig_names_2 <- names(df.2)
df.2 <- df.2 %>% clean_names()
response_labels_2 <- setNames(orig_names_2, names(df.2))
df.2 <- df.2 %>% mutate(depth = fct_recode(depth, "0-75" = "0 - 7.5", "75-150" = "7.5 - 15", "150-250" = "15 - 25"))
cols <- c("plot", "depth", "treatment", "rep")
df.2[, cols] <- lapply(df.2[, cols], as.factor)
if (!"months" %in% names(df.2) && "date" %in% names(df.2)) {
  d <- tryCatch(as.Date(df.2$date), error = function(e) NULL)
  if (is.null(d) || all(is.na(d))) d <- as.numeric(df.2$date)
  df.2$months <- as.numeric(difftime(d, min(d, na.rm = TRUE), units = "days")) / (365.25 / 12)
}
df.2 %>% group_by(date, treatment, depth) %>% tally()
```

```{r}
#| label: wrangle-df4

# Microbial biomass C & N: read, clean names, depth labels, factors
df.4 <- read_excel(here("Analysis/Data/compiled_data_for_paper.xlsx"), sheet = "MB C & N")
orig_names_4 <- names(df.4)
df.4 <- df.4 %>% clean_names()
response_labels_4 <- setNames(orig_names_4, names(df.4))
response_labels_4["mbc_mg_kg"]         <- "Microbial biomass C (kg/ha)"
response_labels_4["mbn_mg_kg"]         <- "Microbial biomass N (kg/ha)"
response_labels_4["mbc_kg_ha"]         <- "Microbial biomass C (kg/ha)"
response_labels_4["mbc_0_25_cm_kg_ha"] <- "Microbial biomass C (kg/ha)"
response_labels_4["mbn_kg_ha"]         <- "Microbial biomass N (kg/ha)"
response_labels_4["mbn_0_25_cm_kg_ha"] <- "Microbial biomass N (kg/ha)"
df.4 <- df.4 %>% mutate(depth = fct_recode(depth, "0-75" = "0 - 7.5", "75-150" = "7.5 - 15", "150-250" = "15 - 25"))
cols <- c("plot", "depth", "treatment", "rep")
df.4[, cols] <- lapply(df.4[, cols], as.factor)
df.4 %>% group_by(date, treatment, depth) %>% tally()
```

```{r}
#| label: wrangle-df3

# Total cations & pH: read (skip 2 rows), clean names, depth labels, factors
df.3 <- read_excel(here("Analysis/Data/compiled_data_for_paper.xlsx"), sheet = "Total cations & pH", skip = 2)
orig_names_3 <- names(df.3)
df.3 <- df.3 %>% clean_names()
response_labels_3 <- setNames(orig_names_3, names(df.3))
response_labels_3["p_h"]                    <- "pH"
response_labels_3["cations_0_25_cm_keq_ha"] <- "Total cations remaining (keq/ha)"
response_labels_3["ca_0_25_cm_keq_ha"]     <- "Cations remaining (keq/ha)"
df.3 <- df.3 %>% mutate(depth = fct_recode(depth, "0-75" = "0 - 7.5", "75-150" = "7.5 - 15", "150-250" = "15 - 25"))
cols <- c("plot", "depth", "treatment", "replicate")
df.3[, cols] <- lapply(df.3[, cols], as.factor)
df.3 %>% group_by(year, treatment, depth) %>% tally()
```

```{r}
#| label: wrangle-df5

# N Leached: read range AK8:AT20 (10 cols: year, N leached × 2, cum N leached × 2, drainage × 2, cum drainage × 2, winter rainfall)
df.5 <- read_excel(here("Analysis/Data/compiled_data_for_paper.xlsx"), sheet = "N Leached", range = "AK8:AT20")
names(df.5) <- c(
  "year", "fallow_n_leached", "pasture_n_leached",
  "cummulative_n_leached_fallow", "cummulative_n_leached_pasture",
  "drainage_fallow", "drainage_pasture",
  "Cummulative_drainage_fallow", "Cummulative_drainage_pasture",
  "winter_rainfall"
)

# Pivot to long: one row per year × treatment (Fallow/Pasture)
df.5 <- df.5 %>%
  pivot_longer(
    cols = -c(year, winter_rainfall),
    names_to = "measure",
    values_to = "value"
  ) %>%
  mutate(
    treatment = if_else(grepl("^fallow_|_fallow$", measure), "Fallow", "Pasture"),
    measure = case_when(
      measure %in% c("fallow_n_leached", "pasture_n_leached") ~ "n_leached_",
      measure %in% c("cummulative_n_leached_fallow", "cummulative_n_leached_pasture") ~ "n_leached_cummulative",
      measure %in% c("drainage_fallow", "drainage_pasture") ~ "drainage",
      measure %in% c("Cummulative_drainage_fallow", "Cummulative_drainage_pasture") ~ "cummulative_drainage",
      .default = measure
    )
  ) %>%
  pivot_wider(names_from = measure, values_from = value)

# Factor treatment, drop rows with any NA
df.5 <- df.5 %>%
  mutate(treatment = factor(treatment, levels = c("Fallow", "Pasture"))) %>%
  na.omit()

# Labels for plotting (used if this table is combined with other analyses)
response_labels_5 <- c(
  n_leached_cummulative = "Cumulative N leached (kg N/ha)",
  cummulative_drainage  = "Cumulative drainage (mm)"
)
df.5
```
:::

```{r}
#| label: functions
#| results: false

source(here::here("Scripts", "r", "functions.R"))
```

```{r}
#| label: plots-output-config
#| results: false

# Output directories at project root (next to each other): Plots and Tables
plots_dir         <- here::here("Plots")
tables_dir        <- here::here("Tables")
save_width_plots  <- 5
save_height_plots <- 3
save_dpi_plots    <- 600
```

## Analyses
### Total soil C, N and bulk density

```{r}
#| label: analysis-1-fit
#| results: false

analysis_1 <- fit_models(
  df,
  response_vars = c("tn_g_kg_soil", "tc_g_kg_soil", "c_n_ratio", "bd_g_cm3", "c_stock_t_ha", 
  "n_stock_to_25_cm_kg_ha","c_stock_to_25_cm_t_ha"),
  time = "months",
  treatment = "treatment",
  depth = "depth"
)
```

**Plots**

```{r}
#| label: analysis-1-plots
#| fig-width: 10
#| fig-height: 20
#| out-width: 25%

plot_model_fits(
  model_fits      = analysis_1,
  time            = "months",
  treatment       = "treatment",
  depth           = "depth",
  depth_order     = c("0-75", "75-150", "150-250"),
  raw_data        = df,
  response_labels = response_labels_1,
  ncol            = 2,
  save_dir        = plots_dir,
  save_dpi        = save_dpi_plots,
  save_width      = save_width_plots,
  save_height     = save_height_plots
)
```

**Total soil N (kg/ha) — Fallow only**

```{r}
#| label: analysis-1-plot-tn-fallow
#| fig-width: 10
#| fig-height: 5
#| out-width: 25%

plot_model_fits(
  model_fits        = analysis_1,
  time              = "months",
  treatment         = "treatment",
  depth             = "depth",
  depth_order       = c("0-75", "75-150", "150-250"),
  raw_data          = df,
  response_labels   = response_labels_1,
  response_filter   = "tn_g_kg_soil",
  treatment_filter  = "Fallow",
  ncol              = 2,
  save_dir          = plots_dir,
  save_prefix       = "tn_fallow",
  save_dpi          = save_dpi_plots,
  save_width        = save_width_plots,
  save_height       = save_height_plots
)
```

**AICc table**

::: {.aicc-table}
```{r}
#| label: analysis-1-aicc

knitr::kable(analysis_1$aicc_table)
```
:::

::: {.callout-note title="Model checks (click to show/hide)" collapse="true"}
```{r}
#| label: analysis-1-checks
#| eval: true
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

check_model_performance(analysis_1)
```
:::

::: {.callout-note title="Best model summaries (click to show/hide)" collapse="true"}
::: {.model-summary}
```{r}
#| label: analysis-1-summaries
#| class-output: model-summary

best_model_summaries(analysis_1)
```
:::
:::

### Particulate organic matter (POM) C and N

```{r}
#| label: analysis-2-fit
#| results: false

analysis_2 <- fit_models(
  df.2,
  response_vars = c("soil_mass_0_25_cm_t_ha", "total_pom_c_mg_ha", "total_pom_n_kg_ha"),
  time = "months",
  treatment = "treatment",
  depth = "depth"
)
```

**Plots**

```{r}
#| label: analysis-2-plots
#| fig-width: 10
#| fig-height: 10
#| out-width: 25%

plot_model_fits(
  model_fits      = analysis_2,
  time            = "months",
  treatment       = "treatment",
  depth           = "depth",
  depth_order     = c("0-75", "75-150", "150-250"),
  raw_data        = df.2,
  response_labels = response_labels_2,
  ncol            = 2,
  save_dir        = plots_dir,
  save_dpi        = save_dpi_plots,
  save_width      = save_width_plots,
  save_height     = save_height_plots
)
```

**AICc table**

::: {.aicc-table}
```{r}
#| label: analysis-2-aicc

knitr::kable(analysis_2$aicc_table)
```
:::

::: {.callout-note title="Model checks (click to show/hide)" collapse="true"}
```{r}
#| label: analysis-2-checks
#| eval: true
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

check_model_performance(analysis_2)
```
:::

::: {.callout-note title="Best model summaries (click to show/hide)" collapse="true"}
::: {.model-summary}
```{r}
#| label: analysis-2-summaries
#| class-output: model-summary

best_model_summaries(analysis_2)
```
:::
:::

### Microbial biomass C and N

```{r}
#| label: analysis-3-fit
#| results: false

analysis_3 <- fit_models(
  df.4,
  response_vars = c("mbc_mg_kg", "mbn_mg_kg", "mbc_kg_ha", "mbc_0_25_cm_kg_ha", "mbn_kg_ha", "mbn_0_25_cm_kg_ha"),
  time = "months",
  treatment = "treatment",
  depth = "depth"
)
```

**Plots**

```{r}
#| label: analysis-3-plots
#| fig-width: 10
#| fig-height: 15
#| out-width: 25%

plot_model_fits(
  model_fits      = analysis_3,
  time            = "months",
  treatment       = "treatment",
  depth           = "depth",
  depth_order     = c("0-75", "75-150", "150-250"),
  raw_data        = df.4,
  response_labels = response_labels_4,
  ncol            = 2,
  save_dir        = plots_dir,
  save_dpi        = save_dpi_plots,
  save_width      = save_width_plots,
  save_height     = save_height_plots
)
```

**AICc table**

::: {.aicc-table}
```{r}
#| label: analysis-3-aicc

knitr::kable(analysis_3$aicc_table)
```
:::

::: {.callout-note title="Model checks (click to show/hide)" collapse="true"}
```{r}
#| label: analysis-3-checks
#| eval: true
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

check_model_performance(analysis_3)
```
:::

::: {.callout-note title="Best model summaries (click to show/hide)" collapse="true"}
::: {.model-summary}
```{r}
#| label: analysis-3-summaries
#| class-output: model-summary

best_model_summaries(analysis_3)
```
:::
:::

### Soil pH and cations

```{r}
#| label: analysis-4-fit
#| results: false

analysis_4 <- fit_models(
  df.3,
  response_vars = c("p_h", "cations_0_25_cm_keq_ha", "ca_0_25_cm_keq_ha"),
  time = "months",
  treatment = "treatment",
  depth = "depth"
)
```

**Plots**

```{r}
#| label: analysis-4-plots
#| fig-width: 10
#| fig-height: 10
#| out-width: 25%

plot_model_fits(
  model_fits      = analysis_4,
  time            = "months",
  treatment       = "treatment",
  depth           = "depth",
  depth_order     = c("0-75", "75-150", "150-250"),
  raw_data        = df.3,
  response_labels = response_labels_3,
  ncol            = 2,
  save_dir        = plots_dir,
  save_dpi        = save_dpi_plots,
  save_width      = save_width_plots,
  save_height     = save_height_plots
)
```

**AICc table**

::: {.aicc-table}
```{r}
#| label: analysis-4-aicc

knitr::kable(analysis_4$aicc_table)
```
:::

**Export AICc tables to Excel**

```{r}
#| label: export-aicc-tables
#| echo: false

dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
writexl::write_xlsx(
  list(
    Analysis_1_Total_soil_C_N   = analysis_1$aicc_table,
    Analysis_2_POM_C_N          = analysis_2$aicc_table,
    Analysis_3_Microbial_biomass = analysis_3$aicc_table,
    Analysis_4_pH_cations       = analysis_4$aicc_table
  ),
  file.path(tables_dir, "aicc_tables.xlsx")
)
```

::: {.callout-note title="Model checks (click to show/hide)" collapse="true"}
```{r}
#| label: analysis-4-checks
#| eval: true
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

check_model_performance(analysis_4)
```
:::

::: {.callout-note title="Best model summaries (click to show/hide)" collapse="true"}
::: {.model-summary}
```{r}
#| label: analysis-4-summaries
#| class-output: model-summary

best_model_summaries(analysis_4)
```
:::
:::

### N Leached

Linear regression: cumulative N leached ~ cumulative drainage × treatment (`lm`).

```{r}
#| label: analysis-5-fit
#| results: false

# Linear model: cumulative N leached ~ cumulative drainage × treatment (separate slope per treatment)
model_n_leached <- lm(n_leached_cummulative ~ cummulative_drainage * treatment, data = df.5)
```

**Plots**

```{r}
#| label: analysis-5-plots
#| fig-width: 5
#| fig-height: 5
#| out-width: 50%
#| layout-ncol: 2

# Prediction grid: drainage × treatment for fitted lines and 95% CI
pred_n_leached <- expand_grid(
  cummulative_drainage = seq(
    min(df.5$cummulative_drainage, na.rm = TRUE),
    max(df.5$cummulative_drainage, na.rm = TRUE),
    length.out = 100
  ),
  treatment = factor(c("Fallow", "Pasture"), levels = c("Fallow", "Pasture"))
)
pred_n_leached$fit <- predict(model_n_leached, newdata = pred_n_leached, se.fit = TRUE)$fit
pred_n_leached$se  <- predict(model_n_leached, newdata = pred_n_leached, se.fit = TRUE)$se.fit
pred_n_leached     <- pred_n_leached %>% filter(complete.cases(fit, se))

# Subtitle: Best model, R², and equations (match style of other analysis plots)
co         <- coef(model_n_leached)
r2         <- format(round(summary(model_n_leached)$r.squared, 2), trim = TRUE)
int_fallow <- co["(Intercept)"]
slope_fallow <- co["cummulative_drainage"]
int_pasture  <- co["(Intercept)"] + if (is.na(co["treatmentPasture"])) 0 else co["treatmentPasture"]
slope_pasture <- co["cummulative_drainage"] +
  if (is.na(co["cummulative_drainage:treatmentPasture"])) 0 else co["cummulative_drainage:treatmentPasture"]
eq_fallow  <- paste0(
  "Fallow: y = ", format(round(int_fallow, 2), trim = TRUE),
  if (slope_fallow >= 0) " + " else " \u2212 ",
  format(round(abs(slope_fallow), 2), trim = TRUE), "\u00b7Cumulative drainage (mm)"
)
eq_pasture <- paste0(
  "Pasture: y = ", format(round(int_pasture, 2), trim = TRUE),
  if (slope_pasture >= 0) " + " else " \u2212 ",
  format(round(abs(slope_pasture), 2), trim = TRUE), "\u00b7Cumulative drainage (mm)"
)
subtitle_n_leached <- paste("Best model: Linear", paste0("R\u00b2 = ", r2), eq_fallow, eq_pasture, sep = "\n")

# Plot: points, fitted lines, and 95% CI ribbon by treatment
p_n_leached <- ggplot(df.5, aes(x = cummulative_drainage, y = n_leached_cummulative, colour = treatment)) +
  geom_ribbon(
    data = pred_n_leached,
    aes(x = cummulative_drainage, ymin = fit - 1.96 * se, ymax = fit + 1.96 * se, fill = treatment),
    alpha = 0.15, colour = NA, inherit.aes = FALSE
  ) +
  geom_line(
    data = pred_n_leached,
    aes(x = cummulative_drainage, y = fit, colour = treatment),
    linewidth = 1.2, inherit.aes = FALSE
  ) +
  geom_point(alpha = 0.8, size = 2.5) +
  scale_colour_manual(values = c(Fallow = "#8B4513", Pasture = "#228B22"), name = "Treatment") +
  scale_fill_manual(values = c(Fallow = "#8B4513", Pasture = "#228B22"), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.32))) +
  scale_y_continuous(expand = expansion(mult = 0.12)) +
  labs(
    x = "Cumulative drainage (mm)",
    y = "Cumulative N leached (kg N/ha)",
    subtitle = subtitle_n_leached
  ) +
  theme_classic(base_size = 11) +
  theme(
    aspect.ratio = 0.5,
    legend.position = c(1, 0.5),
    legend.justification = c(1, 0.5),
    legend.box = "vertical",
    legend.box.background = element_rect(fill = "white", colour = "grey85", linewidth = 0.5),
    legend.margin = margin(6, 8, 6, 8),
    plot.subtitle = element_text(size = 8, colour = "grey40", hjust = 0, lineheight = 1.1, margin = margin(b = 10))
  )
print(p_n_leached)
# Same workflow as other plots: save to plots_dir with shared dimensions
ggsave(file.path(plots_dir, "n_leached_vs_drainage.png"), plot = p_n_leached + labs(subtitle = NULL) + scale_y_continuous(expand = expansion(mult = 0.096)) + theme(plot.margin = margin(1.6, 2, 1.6, 2, "pt")), width = save_width_plots, height = save_height_plots, dpi = save_dpi_plots)
```

::: {.callout-note title="Model checks (click to show/hide)" collapse="true"}
```{r}
#| label: analysis-5-checks
#| eval: true
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

# Diagnostic plots for lm (normality, linearity, homogeneity, etc.)
print(plot(performance::check_model(model_n_leached, panel = TRUE)))
```
:::

::: {.callout-note title="Model summary and statistical significance" collapse="true"}
::: {.model-summary}
```{r}
#| label: analysis-5-summaries
#| class-output: model-summary

# Coefficients, SE, t, p-value, R²
summary(model_n_leached)
```
:::
:::

# References

R Core Team (2022). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.

::: {.callout-note title="Session Information" collapse="true"}
```{r}
#| label: session-info
#| echo: false

utils::sessionInfo()
```
:::
